An elementary and constructive proof of Grothendieck's generic freeness lemma

### 2018-05-15 paper-generic-freeness.tex
Improved the presentation; fought the index war.

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,tikz}
\usetikzlibrary{patterns,matrix}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{hyperref}

@@

\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\defeq}{\vcentcolon=}
\DeclareMathOperator{\Spec}{Spec}

@@

\maketitle

@@

The following proposition is just a special instance of Grothendieck's generic
freeness lemma. We include it here because we want to present its proof, which
is easier and shorter than the proof of the general case. The general proof
will be self-contained and not refer to this one.

@@

@@

\begin{lemma}\label{lemma:basis}
Let~$A$ be a ring. Let~$M$ be an~$A$-module with generating
family~$(x_i)_{i \in I}$ where~$I$ is a totally ordered set. Assume that the
only element~$g \in A$ such that one of the~$x_i$ is an~$A[g^{-1}]$-linear
combination in~$M[g^{-1}]$ of other generators with smaller index is~$g = 0$.
Then~$M$ is free with~$(x_i)_{i \in I}$ as a basis.
\end{lemma}

\begin{proof}Let~$\sum_i a_i x_i = 0$. Starting with the greatest
index~$i$ which appears in that sum, we see that in~$M[a_i^{-1}]$, the
element~$x_i$ is an~$A[g^{-1}]$-linear combination of other generators with
smaller index. Thus~$a_i = 0$ by assumption.\end{proof}

\begin{proof}[Proof of Theorem~\ref{thm:algebraic}]
Let~$B$ be generated by~$(x_1,\ldots,x_n)$ as an~$A$-algebra and
let~$M$ be generated by~$(v_1,\ldots,v_m)$ as a~$B$-module. We endow the sets
\begin{align*}
  I &\defeq \{ (i_1,\ldots,i_n) \,|\, i_1,\ldots,i_n \in \{ 0,1,\ldots \} \}
  \quad\text{and} \\
  J &\defeq \{ (\ell, i_1,\ldots,i_n) \,|\, \ell \in \{ 1,\ldots,m \}, i_1,\ldots,i_n \in \{ 0,1,\ldots \} \}
\end{align*}
with the lexicographic order. The family~$(w_j)_{j \in J} \defeq (x_1^{i_1}
\cdots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in J}$ generates~$M$ as
an~$A$-module, and we'll call a subfamily~$(w_j)_{j \in J' \subseteq J}$
\emph{good} if and only if for all~$j \in J$, the vector~$w_j$ is a linear
combination of the vectors~$(w_{j'})_{j' \in J, j' \preceq j}$,
and if~$(\ell,i_1,\ldots,i_n) \not\in J'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
I'$ for all~$k_1 \geq i_1, \ldots, k_n \geq i_n$. Figure~\ref{fig:good} shows
how a good generating family can look like.

\begin{figure}[b]
  \begin{tikzpicture}[scale=1.0]
    \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

@@

    % !perl -we 'for my $y (reverse 0..7) { for my $x (0..7) { print % "\$x^$x\Ey^$y\$ & " } print "\\\\ \n" }'
    \begin{scope}[yshift=4cm, xshift=4cm]
      \matrix[matrix of nodes,nodes={inner sep=0pt,text width=1.0cm,align=center,minimum height=1.0cm}]{
        \tiny $x^0y^7v_1$ & \tiny $x^1y^7v_1$ & \tiny $x^2y^7v_1$ & \tiny $x^3y^7v_1$ & \tiny $x^4y^7v_1$ & \tiny $x^5y^7v_1$ & \tiny $x^6y^7v_1$ & \tiny $x^7y^7v_1$ & \\
        \tiny $x^0y^6v_1$ & \tiny $x^1y^6v_1$ & \tiny $x^2y^6v_1$ & \tiny $x^3y^6v_1$ & \tiny $x^4y^6v_1$ & \tiny $x^5y^6v_1$ & \tiny $x^6y^6v_1$ & \tiny $x^7y^6v_1$ & \\
        \tiny $x^0y^5v_1$ & \tiny $x^1y^5v_1$ & \tiny $x^2y^5v_1$ & \tiny $x^3y^5v_1$ & \tiny $x^4y^5v_1$ & \tiny $x^5y^5v_1$ & \tiny $x^6y^5v_1$ & \tiny $x^7y^5v_1$ & \\
        \tiny $x^0y^4v_1$ & \tiny $x^1y^4v_1$ & \tiny $x^2y^4v_1$ & \tiny $x^3y^4v_1$ & \tiny $x^4y^4v_1$ & \tiny $x^5y^4v_1$ & \tiny $x^6y^4v_1$ & \tiny $x^7y^4v_1$ & \\
        \tiny $x^0y^3v_1$ & \tiny $x^1y^3v_1$ & \tiny $x^2y^3v_1$ & \tiny $x^3y^3v_1$ & \tiny $x^4y^3v_1$ & \tiny $x^5y^3v_1$ & \tiny $x^6y^3v_1$ & \tiny $x^7y^3v_1$ & \\
        \tiny $x^0y^2v_1$ & \tiny $x^1y^2v_1$ & \tiny $x^2y^2v_1$ & \tiny $x^3y^2v_1$ & \tiny $x^4y^2v_1$ & \tiny $x^5y^2v_1$ & \tiny $x^6y^2v_1$ & \tiny $x^7y^2v_1$ & \\
        \tiny $x^0y^1v_1$ & \tiny $x^1y^1v_1$ & \tiny $x^2y^1v_1$ & \tiny $x^3y^1v_1$ & \tiny $x^4y^1v_1$ & \tiny $x^5y^1v_1$ & \tiny $x^6y^1v_1$ & \tiny $x^7y^1v_1$ & \\
        \tiny $x^0y^0v_1$ & \tiny $x^1y^0v_1$ & \tiny $x^2y^0v_1$ & \tiny $x^3y^0v_1$ & \tiny $x^4y^0v_1$ & \tiny $x^5y^0v_1$ & \tiny $x^6y^0v_1$ & \tiny $x^7y^0v_1$ & \\
      };
    \end{scope}
  \end{tikzpicture}
  \caption{\label{fig:good}A graphical depiction of a good generating family (the non-hatched cells) in
  the special case~$n = 2, m = 1$, writing~``$x$'' and~``$y$'' for~$x_1$ and~$x_2$. The hatched cells indicate vectors which
  have already been removed from the family. The small black squares indicate
  \emph{corners}. If the vector in the red cell will be found to be expressible
  as a linear combination of vectors with smaller index (blue cells), it will
  be removed, along with the vectors in all cells to the top and to the right
  of the red cell.}
\end{figure}

Similarly, we define when a subfamily of the canonical generating
family~$(x_1^{i_1} \cdots x_n^{i_n})_{(i_1,\ldots,i_n) \in I}$ of~$B$ is good
(which is just the special case~$m = 1$). We then proceed by induction on the
shapes of a given good generating family~$(w_j)_{j \in J'}$ for~$M$ and a given
good generating family~$(s_i)_{i \in I'}$ for~$B$, starting with the canonical
ones.

We show that~$(w_j)_{j \in J'}$ is a basis of~$M$ by verifying the assumptions
of Lemma~\ref{lemma:basis}. Thus let~$g \in A$ be given such that one of
the~$w_j$ is an $A[g^{-1}]$-linear combination of generators with smaller index
in~$M[g^{-1}]$. Removing~$w_j = x_1^{i_1} \cdots x_n^{i_n} v_\ell$ and also all
vectors~$x_1^{k_1} \cdots x_n^{k_n} v_\ell$ where~$k_1 \geq i_1, \ldots,
k_n \geq i_n$, we obtain a subfamily which is still good for
the~$A[g^{-1}]$-module~$M[g^{-1}]$. By the induction hypothesis, applied
to~$A[g^{-1}]$ and its module~$M[g^{-1}]$, it therefore follows that $A[g^{-1}] = 0$.
This implies that $g = 0$ since~$A$ is reduced.

Similarly, we show that the given good generating family~$(s_i)_{i \in I'}$ is
a basis. Thus~$M$ and~$B$ are free over~$A$. We fix for any corner~$j =
(\ell,i_1,\ldots,i_n) \in J$, as indicated in Figure~\ref{fig:good},
a way of expressing~$x^{i_1} \cdots x^{i_n} v_\ell = \sum_{(p,k_1,\ldots,k_n)}
a_{j; p,k_1,\ldots,k_n} x^{k_1} \cdots x^{k_n} v_p$ as a linear combination of
generators with strictly smaller index. Then~$M$ is isomorphic to
the~$B$-module
\[ B\langle V_1,\ldots, V_m\rangle/
  (x^{i_1} \cdots x^{i_n} V_\ell - \textstyle\sum_{(p,k_1,\ldots,k_n)}
  a_{j; p,k_1,\ldots,k_n} x_{k_1} \cdots x_{k_n} V_p)_j,
\]
where the index~$j$ ranges over all corners of~$J$, and is therefore finitely
presented. In a similar vain, a quotient algebra of~$A[X_1,\ldots,X_n]$, where
we mod out a suitable ideal with as many generators as corners of~$I'$, is
isomorphic to~$B$.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,tikz,float,caption}
\usetikzlibrary{patterns,matrix}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{hyperref}

@@

\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\defeq}{\vcentcolon=}
\DeclareMathOperator{\Spec}{Spec}

@@

\maketitle
\thispagestyle{empty}

@@

The following proposition is just a special instance of Grothendieck's generic
freeness lemma. Its proof is easier and shorter than the proof of the general
case, which is why we present it here. The general proof
will not refer to this one.

@@

We remark that proof takes a somewhat curious course: Our goal is to
verify~$1 = 0$, but as an intermediate step we verify that~$M$ is free, which
after the fact will be a trivial statement.

@@

\begin{proof}[Proof of Theorem~\ref{thm:algebraic}]
Let~$B$ be generated by~$(x_1,\ldots,x_n)$ as an~$A$-algebra and
let~$M$ be generated by~$(v_1,\ldots,v_m)$ as a~$B$-module. We endow the sets
\begin{align*}
  \I &\defeq \{ (i_1,\ldots,i_n) \,|\, i_1,\ldots,i_n \geq 0 \}
  \quad\text{and} \\
  \J &\defeq \{ (\ell, i_1,\ldots,i_n) \,|\, 1 \leq \ell \leq m, i_1,\ldots,i_n \geq 0 \}
\end{align*}
with the lexicographic order. The family~$(w_J)_{J \in \J} \defeq (x_1^{i_1}
\cdots x_n^{i_n} v_\ell)_{(\ell,i_1,\ldots,i_n) \in \J}$ generates~$M$ as
an~$A$-module, and we'll call a subfamily~$(w_J)_{J \in \J' \subseteq \J}$
\emph{good} if and only if for all~$J \in \J$, the vector~$w_J$ is a linear
combination of the vectors~$(w_{J'})_{J' \in \J', J' \preceq J}$,
and if~$(\ell,i_1,\ldots,i_n) \not\in \J'$ implies~$(\ell,k_1,\ldots,k_n) \not\in
\J'$ for all~$k_1 \geq i_1, \ldots, k_n \geq i_n$. Figure~\ref{fig:good} shows
how a good generating family can look like.
Similarly, we define when a subfamily of the canonical generating
family~$(x_1^{i_1} \cdots x_n^{i_n})_{(i_1,\ldots,i_n) \in \I}$ of~$B$ is good
(which is just the special case~$m = 1$).

\begin{figure}[t]
  \captionsetup{width=.9\linewidth}
  \begin{tikzpicture}[scale=0.9]
    \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

@@

    % !perl -we 'for my $y (reverse 0..7) { for my $x (0..7) { print % "\$x^$x\Ey^$y\$ & " } print "\\\\ \n" }'
    \begin{scope}[yshift=4cm, xshift=4cm]
      \matrix[matrix of nodes,nodes={inner sep=0pt,text width=0.9cm,align=center,minimum height=0.9cm}]{
        \tiny $x^0y^7v_1$ & \tiny $x^1y^7v_1$ & \tiny $x^2y^7v_1$ & \tiny $x^3y^7v_1$ & \tiny $x^4y^7v_1$ & \tiny $x^5y^7v_1$ & \tiny $x^6y^7v_1$ & \tiny $x^7y^7v_1$ & \\
        \tiny $x^0y^6v_1$ & \tiny $x^1y^6v_1$ & \tiny $x^2y^6v_1$ & \tiny $x^3y^6v_1$ & \tiny $x^4y^6v_1$ & \tiny $x^5y^6v_1$ & \tiny $x^6y^6v_1$ & \tiny $x^7y^6v_1$ & \\
        \tiny $x^0y^5v_1$ & \tiny $x^1y^5v_1$ & \tiny $x^2y^5v_1$ & \tiny $x^3y^5v_1$ & \tiny $x^4y^5v_1$ & \tiny $x^5y^5v_1$ & \tiny $x^6y^5v_1$ & \tiny $x^7y^5v_1$ & \\
        \tiny $x^0y^4v_1$ & \tiny $x^1y^4v_1$ & \tiny $x^2y^4v_1$ & \tiny $x^3y^4v_1$ & \tiny $x^4y^4v_1$ & \tiny $x^5y^4v_1$ & \tiny $x^6y^4v_1$ & \tiny $x^7y^4v_1$ & \\
        \tiny $x^0y^3v_1$ & \tiny $x^1y^3v_1$ & \tiny $x^2y^3v_1$ & \tiny $x^3y^3v_1$ & \tiny $x^4y^3v_1$ & \tiny $x^5y^3v_1$ & \tiny $x^6y^3v_1$ & \tiny $x^7y^3v_1$ & \\
        \tiny $x^0y^2v_1$ & \tiny $x^1y^2v_1$ & \tiny $x^2y^2v_1$ & \tiny $x^3y^2v_1$ & \tiny $x^4y^2v_1$ & \tiny $x^5y^2v_1$ & \tiny $x^6y^2v_1$ & \tiny $x^7y^2v_1$ & \\
        \tiny $x^0y^1v_1$ & \tiny $x^1y^1v_1$ & \tiny $x^2y^1v_1$ & \tiny $x^3y^1v_1$ & \tiny $x^4y^1v_1$ & \tiny $x^5y^1v_1$ & \tiny $x^6y^1v_1$ & \tiny $x^7y^1v_1$ & \\
        \tiny $x^0y^0v_1$ & \tiny $x^1y^0v_1$ & \tiny $x^2y^0v_1$ & \tiny $x^3y^0v_1$ & \tiny $x^4y^0v_1$ & \tiny $x^5y^0v_1$ & \tiny $x^6y^0v_1$ & \tiny $x^7y^0v_1$ & \\
      };
    \end{scope}
  \end{tikzpicture}
  \caption{\label{fig:good}A graphical depiction of a good generating family (the non-hatched cells) in
  the special case~$n = 2, m = 1$, writing~``$x$'' and~``$y$'' for~$x_1$ and~$x_2$. The hatched cells indicate vectors which
  have already been removed from the family. The small black squares indicate
  \emph{corners}. If the vector in the red cell will be found to be expressible
  as a linear combination of vectors with smaller index (blue cells), it will
  be removed, along with the vectors in all cells to the top and to the right
  of the red cell.}
\end{figure}

We then proceed by induction on the
shapes of a given good generating family~$(w_J)_{J \in \J'}$ for~$M$ and a given
good generating family~$(s_I)_{I \in \I'}$ for~$B$, starting with the canonical ones. It's
reasonably obvious that this induction is well-founded; the formal statement
that it is so is known as \emph{Dickson's Lemma} in mathematical logic.

We show that~$(w_J)_{J \in \J'}$ is a basis of~$M$ by verifying linear
independence. Thus let~$\sum_J a_J w_J = 0$ in~$M$. We show that all
coefficients in this sum are zero, starting with the largest
appearing index~$J$: In the module~$M[a_J^{-1}]$ over the localized ring~$A[a_J^{-1}]$,
the vector~$w_J$ is a linear combination of generators with
smaller index. Removing~$w_J = x_1^{i_1} \cdots x_n^{i_n} v_\ell$ and also all
vectors~$x_1^{k_1} \cdots x_n^{k_n} v_\ell$ where~$k_1 \geq i_1, \ldots,
k_n \geq i_n$, we obtain a subfamily which is still good for
the localized module. The induction hypothesis, applied
to~$A[a_J^{-1}]$ and its module~$M[a_J^{-1}]$, therefore implies that $A[a_J^{-1}] = 0$.
Thus $a_J = 0$ since~$A$ is reduced.
\enlargethispage{2em}

Similarly, we show that the given good generating family~$(s_I)_{I \in \I'}$ is
a basis. Thus~$M$ and~$B$ are free over~$A$. We fix for any corner~$J$
of~$\J'$, as indicated in Figure~\ref{fig:good},
a way of expressing~$w_J = \sum_K a_{JK} w_K$ as a linear combination of
generators with strictly smaller index. Let~$\widehat w_{(\ell,i_1,\ldots,i_n)}
\defeq x_1^{i_1} \cdots x_n^{i_n} V_\ell$ in the free~$B$-module~$B\langle
V_1,\ldots, V_m\rangle$. The canonical map
\[ \widehat M \defeq B\langle V_1,\ldots, V_m\rangle/
  (\widehat w_J - \textstyle\sum_K a_{JK} \widehat w_K)_{\text{$J$ corner of~$\J'$}} \longrightarrow M
\]
is trivially well-defined and surjective. It is also injective, since any
element of~$\widehat M$ can be written as an~$A$-linear combination of the
vectors~$(\widehat w_J)_{J \in \J'}$ by employing the corner relations a finite
number of times.  Therefore~$M$ is finitely presented as a~$B$-module.

In a similar vein, a quotient algebra of~$A[X_1,\ldots,X_n]$, where
we mod out by a suitable ideal with as many generators as corners of~$\I'$, is
isomorphic to~$B$. Thus~$B$ is finitely presented as an~$A$-algebra.
